Late Stage Review

### 2020-01-08 papers/lsr/lsr.tex
LSR draft (should be about 30 pages, hence ive removed the proofs and other bits)

%\bibliographystyle{apalike}

@@

\usepackage[square,numbers]{natbib}
\usepackage{url}
%\bibliographystyle{elsarticle-harv}
\bibliographystyle{plain}

@@

%\usepackage{media9}
%\graphicspath{ {../sphere/plots} }

\title{Sparse spectral and $p$-finite element methods for partial differential equations on disk slices and trapeziums}
\author{Sheehan Olver, Ben Snowball}
%\date{}							% Activate to display a given date or no date

@@

\begin{abstract}
Sparse spectral methods for solving partial differential equations have been derived in recent years using hierarchies of classical orthogonal polynomials on intervals, disks, and triangles. In this work we extend this methodology to a hierarchy of non-classical orthogonal polynomials on \bstodo{Reviewer 1, comment 3}disk slices and trapeziums. This builds on the observation that sparsity is guaranteed due to the boundary being defined by an algebraic curve, and that the entries of partial differential operators can be determined using formulae in terms of (non-classical) univariate orthogonal polynomials. We apply the framework to solving the Poisson, variable coefficient Helmholtz, and Biharmonic equations. \bstodo{Reviewer 1, comment 3}In this paper we focus on constant Dirichlet boundary conditions, as well as zero Dirichlet and Neumann boundary conditions, with other types of boundary conditions requiring future work.
 \end{abstract}

%
\section{Introduction}

This paper develops sparse spectral methods for solving linear partial differential equations on a special class of geometries that includes disk slices and trapeziums.  
More precisely, we consider the solution of partial differential equations on the domain
\begin{align*}
	\Omega := \{(x,y) \in \R^2 \quad | \quad \alpha < x < \beta, \: \gamma \rho(x) < y < \delta \rho(x)\}
\end{align*}
where  either of the following conditions hold:

@@

For simplicity of presentation we focus on the disk-slice, where $\rho(x) = \sqrt{1-x^2}$, $(\alpha,\beta) \subset (0, 1)$, and $(\gamma, \delta)  = (-1,1)$, and discuss \bstodo{typo 1}an extension to other geometries in the appendix (including the half-disk and trapeziums).

We show that partial differential equations become sparse linear systems when viewed as acting on expansions involving a family of orthogonal polynomials (OPs) that  generalise Jacobi polynomials, mirroring the ultraspherical spectral method for ordinary differential equations \cite{olver2013fast} and its analogue on the triangle \cite{olver2018recurrence,olver2019triangle}.  On the disk-slice the family of weights we consider are of the form
$$
\Wabc(x,y) = (\beta - x)^a \: (x - \alpha)^b \: (1-x^2-y^2)^c, \qfor \alpha \leq x \leq \beta,\quad -\rho(x) \leq y \leq \rho(x).
$$
The corresponding OPs denoted $\hdopnkab(x,y)$, where $n$ denotes the polynomial degree, and $0 \le k \le n$. We define these to be orthogonalised lexicographically, that is,
$$
\hdopnkab(x,y) = C_{n,k} x^{n-k} y^k + (\hbox{lower order terms})
$$
where $C_{n,k} \neq 0$ and ``lower order terms'' can  include degree $n$ polynomials with fewer $y$ terms. The precise normalization arises from their definition in terms of one-dimensional OPs in Definition~\ref{def:OPconstruction}.

Sparsity comes from expanding the domain and range of an operator using different choices of $a$ and $b$. Whereas the sparsity pattern and entries derived in \cite{olver2018recurrence,olver2019triangle} for equations on the triangle and \cite{vasil2016tensor} for equations on the disk results from manipulations of Jacobi polynomials, in the present work we use a more general integration-by-parts argument to deduce the sparsity structure, alongside careful use of the Christoffel-Darboux formula \cite[18.2.2]{DLMF} and quadrature rules to determine the entries. In particular, by exploiting the connection with one-dimensional orthogonal polynomials \bstodo{typo 2}we can construct discretizations of general partial differential operators of size $p(p-1)/2 \times p(p-1)/2$ in $O(p^3)$ operations, where $p$ is the total polynomial degree. This compares favourably to $O(p^6)$ operations if one proceeds na\"ively. Furthermore, we use this framework to derive sparse $p$-finite element methods that are analogous to those of Beuchler and Sch\"oberl on tetrahedra \cite{beuchler2006new}, see also work by Li and Shen \cite{li2010optimal}.

%An alternative approach is to use \sotodoinline{Compare with "Discontinuous Galerkin scheme for the spherical shallow water equations with applications to tsunami modeling and prediction"}

Here is an overview of the paper:

\noindent \secref{OPs}: We present our procedure to gain a (two-parameter) family of 2D orthogonal polynomials (OPs) on the disk-slice domain, by combining 1D OPs on the interval, to form 2D OPs on the disk.

\noindent\secref{PDOs}: We demonstrate that these families will lead to sparse operators, including Jacobi operators representing multiplication by $x$ and $y$, and partial differential operators. We present a method involving use of the Christoffel-Darboux formula \cite[18.2.2]{DLMF} to obtain the recurrence coefficients for the non-classical 1D OPs, allowing us to exactly use the 1D quadrature rules we present to calculate the non-zero entries to the sparse operators.

\noindent\secref{Computation}: We discuss computational issues, in particular, how to realise the results of the preceding sections in practice. We present a method for explicitly deriving the recurrence coefficients of the non-classical 1D OPs we detail in \secref{OPs}. We derive a quadrature rule on the disk-slice that can be used to expand a function in the OP basis up to a given order. Further, we implement function evaluation using the coefficients of the expansion of a given function using the Clenshaw algorithm.

\noindent\secref{Examples}: We demonstrate the proposed technique for solving Poisson, Helmholtz, and Biharmonic equations on the disk-slice.

\noindent\appref{PFEM}: We use the procedure to construct sparse $p$-finite element methods. This lays the groundwork for a future $hp$-finite element method in a disk, where the elements capture the circular geometry precisely.

\noindent\appref{HalfDisk}: We discuss extension to the special case of the end-disk-slice.

\noindent\appref{Trapezium}: We discuss extension to trapezia.

\section{Orthogonal polynomials on the disk-slice and the trapezium}\label{Section:OPs}

In this section we outline the construction and some basic properties of $\hdopnkabcd(x,y)$. The symmetry in the weight allows us to express the polynomials in terms of 1D OPs, and deduce certain properties such as recurrence relationships.

@@

We can express the three-term recurrences associated \bstodo{typo 3}with $\genjac_n^{(a,b,c)}$ and $\jac_n^{(d,c)}$ as 
\begin{align}
	x \genjac_n^{(a,b,c)}(x) &= \beta_n^{(a,b,c)} \genjac_{n+1}^{(a,b,c)}(x) + \alpha_n^{(a,b,c)} \genjac_n^{(a,b,c)}(x) + \beta_{n-1}^{(a,b,c)} \genjac_{n-1}^{(a,b,c)}(x) \label{eqn:Hrecurrence} \\
	y \jac_n^{(d,c)}(y) &= \delta_n^{(d,c)} \jac_{n+1}^{(d,c)}(y) + \gamma_{n}^{(d,c)} \jac_{n}^{(d,c)}(y) + \delta_{n-1}^{(d,c)} \jac_{n-1}^{(d,c)}(y). \label{eqn:Precurrence}
\end{align}
Of course, for the disk-slice case, we have that $c = d$ and $\gamma_{n}^{(c,c)} = 0 \quad \forall n=0,1,2,\dots$. We can use (\ref{eqn:Hrecurrence}) and (\ref{eqn:Precurrence}) to determine the 2D recurrences for $\hdopnkabcd(x,y)$. Importantly, we can deduce sparsity in the recurrence relationships:

@@

%
\section{Sparse partial differential operators}\label{Section:PDOs}

@@

\begin{proof}
First, note that:
\begin{align}
	\genjacw^{(a,b,c) \: \prime}(x) &= - a \: \genjacw^{(a-1,b,c)}(x) + b \: \genjacw^{(a,b-1,c)}(x) + c \: \rho(x) \: \rho'(x) \:\genjacw^{(a,b,c-2)}(x), \label{eqn:derivativeofweightgenjac} \\
	\jacw^{(c) \: \prime}(y) &= - 2c \: y \: \jacw^{(c-1)}(y), \label{eqn:derivativeofweightjac} \\
	\rho(x) \: \rho'(x) &= -x. \label{eqn:rhoderivative}
\end{align}

>>>>

@@

\usepackage{natbib}
\usepackage{url}
%\bibliographystyle{elsarticle-harv}
\bibliographystyle{plain}

@@

\title{Late Stage Review}
\author{Ben Snowball (Principle Supervisor: Sheehan Olver)}
\date{}							% Activate to display a given date or no date

@@

\section{Introduction}

The goal of my research is to eventually investigate sparse spectral methods on triangulations of the sphere as an alternative to the spherical harmonics approach currently in use at the European Centre for Medium-range Weather Forecasts (ECMWF) in their weather and climate model \cite{wedi2013fast}. The aim would be to preserve the predictive skill of the current approach whilst avoiding the parallel scalability bottleneck from the global spectral transform, which is expected to inhibit the future performance of the ECMWF model.

This document is to provide a brief outline of my work to date, with a closer look at our most recent area of research (solving PDEs on the half disk using sparse spectral methods). The work outlined in this report serves as initial steps to pave the way to solving PDEs on spherical triangles. All code used thus far and from hence forth will mainly be written in the Julia programming language.

\section{Professional Skills Development Requirement}
I have met the initial part of the professional skills development requirement which is to attend a minimum of 2 further professional development workshops by the Late Stage Review (LSR). The 3 further courses taken are: 
\begin{itemize}
	\item Maximising Management Skills 1: Becoming an Effective Researcher
	\item Maximising Management Skills 2: Time Management Strategies for your PhD
	\item Maximising Management Skills 4: Put Project Management into Action
\end{itemize}
The 100 hours of PG courses/activities have been met by virtue of the courses taken in the first year of my studies with the Mathematics of Planet Earth CDT.

\section{Orthogonal polynomials on the sphere (spherical harmonics)}

The first step was to develop a small code library so that we can use spherical harmonics as orthogonal polynomials in three variables $x, y, z$  to evaluate functions on the sphere.

We can the define the spherical harmonics, orthogonal on the unit sphere as \cite{efthimiou2014spherical, dai2013approximation}, \cite[14.30.1, 14.3.1, 14.3.6]{DLMF}:
\begin{align}
\Ylmfull &:= \clm \eimphi \Plm (\costheta) \\
&= \alphalm (1 - (\costheta)^2)^\frac{|m|}{2} \eimphi P^{(|m|,|m|)}_{l-|m|}(\costheta), \quad \text{where } 0 \le |m| \le l, \, l \in \No.
\end{align} 
Note that we can then express \(\Ylm\) in terms of \(x,y,z\) instead of \(\theta, \varphi\) by noting that \(\costheta = z\) and that \( \eimphi\) can be expressed in terms of \(x,y,z\) for any \(m\in\mathbb{Z}\). Indeed, they are polynomials in $x,y,z$ which we denote $\Ylm(x,y,z)$. They span all polynomials modulo the ideal generated by $x^2 + y^2 + z^2 - 1$. While this is equivalent to standard techniques in the litrature concerning spherical harmonics, we approach this with our goal in mind of developing sparse spectral methods on spherical triangles. Thus, we try to use language and formulations that will translate as we move through our steps towards that goal.

\citep{dunkl2014orthogonal} (p75-85) provided a framework for gaining a recurrence relation for each \(\bigP_{l+1}\) in terms of the previous two sub-vectors \(\Pl\) and \(\bigP_{l-1}\), utilising the sub matrices that make up the block-tridiagnonal Jacobi operator matrices for multiplication by $x, y$ and $z$. Clenshaw's algorithm then gives us a way of evaluating a function given its expansion in the spherical harmonic basis (up to a finite order). This provides an optimal complexity method for evaluating spherical harmonic expansions; an alternative to using the explicit expressions that may have benefits, though further investigation is required.

%\subsection{Deriving expressions for the multiplication by \(x, y, z\) of the spherical harmonics}
%
%We start by expressing \(x\,\Ylm(x,y,z)\), \(y\,\Ylm(x,y,z)\) , and \(z\,\Ylm(x,y,z)\)  in terms of \(Y^{m'}_{l'}(x,y,z)\) for any point \((x,y,z)\) on the unit circle.
%
%For \(l \in \No\), \(m \in \Z\) s.t. \(0 \le |m| \le l\):
%\begin{align}
%x\,\Ylm(x,y,z) &= \Alm Y^{m+1}_{l+1}(x,y,z) +  \Blm Y^{m+1}_{l-1}(x,y,z) \nonumber \\
%		     & \quad \quad \quad + \Dlm Y^{m-1}_{l+1}(x,y,z) + \Elm Y^{m-1}_{l-1}(x,y,z), \\
%y\,\Ylm(x,y,z) &= - i \, \big[\Alm Y^{m+1}_{l+1}(x,y,z) +  \Blm Y^{m+1}_{l-1}(x,y,z) \big] \nonumber \\
%		     &\quad \quad \quad + i \, \big[ \Dlm Y^{m-1}_{l+1}(x,y,z) + \Elm Y^{m-1}_{l-1}(x,y,z) \big], \\
%z\,\Ylm(x,y,z) &= \Flm Y^{m}_{l+1}(x,y,z) + \Glm Y^{m}_{l-1}(x,y,z) ,
%\end{align}
%where \(\Alm, \Blm, \Dlm, \Elm, \Flm, \Glm \in \R\) have explicit derivable and calculable formulae.
%
%\subsection{Jacobi matrices}
%
%Define, for \(l \in \No\), \(\Pl\) as the column vector of the degree \(l\) spherical harmonic polynomials, and \(\bigP\) as the stacked block vector of the \(\Pl\)'s; that is
%\begin{align}
%\Pl := \begin{bmatrix}
%		Y^{-l}_l \\
%		\vdots \\
%		Y^l_l
%	\end{bmatrix} \in \C^{2l+1}, 
%\quad \quad 
%\bigP := \begin{bmatrix}
%		\bigP_0 \\
%		\hline
%		\bigP_1 \\
%		\hline
%		\bigP_2 \\
%		\vdots \\
%	\end{bmatrix}.
%\end{align}
%
%Define the (Jacobi) matrices \(J^x, J^y, J^z\) by 
%\begin{align}
%J^x \bigP = x \bigP, \quad J^y \bigP = y \bigP, \quad J^z \bigP = z \bigP.
%\end{align}
%Then, using equations (3--5), we have that the Jacobi matrices each have the following block-tridiagonal form, allowing for easy application computationally:
%\begin{align}
%J^{x/y/z} &= \begin{bmatrix}
%		B^{x/y/z}_0 & A^{x/y/z}_0 & & & & \\
%		C^{x/y/z}_1 & B^{x/y/z}_1 & A^{x/y/z}_1 & & & \\
%		& C^{x/y/z}_2 & B^{x/y/z}_2 & A^{x/y/z}_2  & & & \\
%		& & C^{x/y/z}_3 & \ddots & \ddots & \\
%		& & & \ddots & \ddots & \ddots \\
%	\end{bmatrix}
%\end{align}
%
%\subsection{Three-term recurrence relation for \(\bigP\)}
%
%Combining each system in (7) we can write the tridiagonal-block system
%\begin{align}
%\renewcommand\arraystretch{1.3}
%\mleft[
%\begin{array}{cccc}
%		1  & & & \\
%		B_0-G_0(x,y,z) & A_0 & & \\
%		C_1 & B_1-G_1(x,y,z) & \quad A_1 \quad & \\
%		& C_2 & \ddots & \ddots \\
%		& & \ddots &
%\end{array}
%\mright]
%\bigP
%=
%\begin{bmatrix}
%	\alpha^0_0 \\ 0 \\ 0 \\ \vdots \\ \\
%\end{bmatrix}.
%\end{align}
%where for each \(l \in \No\):
%\begin{align}
%A_l &:= \begin{bmatrix}
%		A^x_l \\
%		A^y_l \\
%		A^z_l
%	    \end{bmatrix} \in \R^{3(2l+1)\times(2l+3)}, \quad
%C_l := \begin{bmatrix}
%		C^x_l \\
%		C^y_l \\
%		C^z_l
%	    \end{bmatrix} \in \R^{3(2l+1)\times(2l-1)} \quad (l \ne 0), \\
%B_l &:= \begin{bmatrix}
%		B^x_l \\
%		B^y_l \\
%		B^z_l
%	    \end{bmatrix} \in \R^{3(2l+1)\times(2l+1)}, \quad
%G_l(x,y,z) := \begin{bmatrix}
%		xI_{2l+1} \\
%		yI_{2l+1} \\
%		zI_{2l+1}
%	    \end{bmatrix} \in \R^{3(2l+1)\times(2l+1)}.
%\end{align}
%
%For each \(l \in \No\) let \(\Dlt\) be any matrix that is a left inverse of \(A_l\), i.e. such that \(\Dlt A_l = I_{2l+3}\). Multiplying our system by the preconditioner matrix that is given by the block diagonal matrix of the \(\Dlt\)'s, we obtain a lower triangular system \citep{dunkl2014orthogonal} (p78), which can be expanded to obtain the recurrence:
%\begin{align}
%\begin{cases}
%\bigP_{-1} := 0 \\
%\bigP_{0} := \alpha^0_0 \\
%\bigP_{l+1} = -\Dlt (B_l-G_l) \Pl - \Dlt C_l  \,\bigP_{l-1}, \quad l \in \N.
%\end{cases}
%\end{align}
%
%\subsection{Evaluation of a scalar function on the sphere}
%We can use a multivariate analogue of the Clenshaw algorithm to evaluate a function at a given point \((x,y,z)\) on the unit sphere provided we know the coefficients of the function when expanded in the spherical harmonic basis, i.e. suppose \(f(x,y,z)\) is a function and we know the set \(\{\bold{f}_l\}\) s.t.
%\begin{align}
%f(x,y,z) \approx \sum^N_{l=0} \bold{f}_l^T \, \Pl (x,y,z), \quad \text{where } \Pl (x,y,z), \bold{f}_l \in \R^{2l+1} \text{ for each } l \in \{0,\dots,N\}.
%\end{align}
%
%The Clenshaw algorithm is then as follows:
%\begin{align*}
%\quad &\text{1) } \text{Set } \bm{\gamma}_{N+2} = \bold{0}, \: \bm{\gamma}_{N+2} = \bold{0}. \\
%\quad &\text{2) } \text{For } n = N:-1:1 \\
%\quad & \quad \quad \quad \text{set } \bm{\gamma}_{n}^T = \bold{f}_n^T - \bm{\gamma}_{n+1}^T D^T_n (B_n - G_n(x,y,z)) -  \bm{\gamma}_{n+2}^T D^T_{n+1}C_{n+1} \\
%\quad &\text{3) } \text{Output: } f(x,y,z) \approx \bigP_0(x,y,z) f_0 + \bm{\gamma}_{1}^T \bigP_1(x,y,z) - \bigP_0(x,y,z) \bm{\gamma}_{2}^T D^T_1 C_1.
%\end{align*}
%where $\bigP_0 \equiv \alpha^0_0$.
%
%We can use the Clenshaw algorithm to output a matrix operator for a function, given we know the coefficients of its expansion in the SH basis. We simply make the adjustment that
%\begin{align*}
%\quad &\text{1) } \text{Set } \bm{\gamma}_{N+2} = \bold{0}, \: \bm{\gamma}_{N+2} = \bold{0}. \\
%\quad &\text{2) } \text{For } n = N:-1:1 \\
%\quad & \quad \quad \quad \text{set } \bm{\gamma}_{n}^T = (\bold{f}_n^T)I - \bm{\gamma}_{n+1}^T D^T_n (B_n - G_n(J^x,J^y,J^z)) -  \bm{\gamma}_{n+2}^T D^T_{n+1}C_{n+1} \\
%\quad &\text{3) } \text{Output: } f(J^x,J^y,J^z) \approx F := \alpha^0_0 f_0 + \bm{\gamma}_{1}^T \bigP_1(J^x,J^y,J^z) - \alpha^0_0 \bm{\gamma}_{2}^T D^T_1 C_1.
%\end{align*}
%where $\bigP_1(J^x,J^y,J^z)$ is a vector of matrices, $G_n(J^x,J^y,J^z)$ is a matrix of matrices, and so each $\bm{\gamma}_n$ is a vector of matrices.

\subsection{Tangent space}

By noting that the tangent space is then spanned by $\{\gradYlm, \gradpYlm\}$, and that these gradients and perpendicular gradients are orthogonal polynomials themselves, we can use the same ideas to create a framework for functions in the tangent space as expansions in this OP basis. Note that the perpendicular gradient is related to the regular surface gradient by
\begin{align}
\gradpYlm(x,y,z) = \unitvec \times \gradYlm(x,y,z),
\end{align}
where \(\unitvec\) is the unit vector normal to the surface of the sphere at the point \((x,y,z)\), i.e. as we are looking at the unit sphere, \(\unitvec\) is simply given by
\begin{align}
\unitvec = \begin{bmatrix} x \\ y \\ z \end{bmatrix}.
\end{align}
For any scalar spherical harmonic (SSH) $\Ylm$, the triad of $\gradYlm, \: \unitvec \times \gradYlm, \: \unitvec \Ylm$ can be referred to as vector spherical harmonics \cite{barrera1985vector}.

Since the spherical harmonics are a basis for the surface of the sphere, and the tangent space of the sphere is spanned by the gradient  and perpendicular gradient of a scalar function, we have that the gradients and perpendicular gradients of the spherical harmonics are a basis for the tangent space, namely \(\gradYlm\), \(\gradpYlm\).

Importantly,  \(\gradYlm\), \(\gradpYlm\) are in fact vector-valued orthogonal polynomials, that span all vector-valued polynomials modulo the vanishing ideal $\{ \bm{p} \in (\Pi_3)^3  : \unitvec \cdot \bm{p} = 0 \}$. A simple calculation shows that such orthogonal polynomials must still have block-tridiagonal Jacobi operators, as multiplication by $x, y$, or $z$ remains inside the ideal. In the following, we determine the coefficients of these Jacobi operators, as well as other important operators needed.

\subsubsection{Spin-1 tensor spherical harmonics}

The gradient and perpendicular gradient of a spherical harmonic \(\Ylm(x,y,z)\) can be expressed in terms of spin-1 tensor spherical harmonics, which in turn can each be expressed as a vector-weighted sum of spherical harmonics.

For any \(l \in \No\), \(m \in \Z\) s.t. \(0 \le |m| \le l\) we have that
\begin{align}
\nabla \Ylm &= \blmone \, \curlyy^{l-1,1}_{l,m} + \blone \curlyy^{l+1,1}_{l,m}, \\
\nabla^\perp \Ylm &= \blzero \, \curlyy^{l,1}_{l,m},
\end{align}
where
\begin{align}
\blmone &:= (l+1) \, \Big(\frac{l}{2l+1}\Big)^\half, \quad
\blzero := i \, \big(l(l+1)\big)^\half, \quad
\blone := l \, \Big(\frac{l+1}{2l+1}\Big)^\half.
\end{align}

We can then gain similar (vector analogue) derivations for recurrences, Jacobi matrices and the Clenshaw algorithm implementation for the tangent space vector-valued orthogonal polynomial basis.

The aim is to, given a PDE say, work with the coefficients of the functions involved (coefficients of their expansion in the OP bases), reframe the PDE in terms of the OP coefficients utilising a suitable timestepping method. For derivative, gradient, divergence etc. operators, we can formulate matrix operators to act on the coefficients, thus making our system to simulate a simple matrix vector system.

\subsection{Example: linear shallow water equations}
Let $\mathbf{u}(x,y,z)$ be the tangential velocity of the flow and $h(x,y,z)$ be the height of the water from some constant reference height $\mathcal{H}$. The linear SWEs are
\begin{align}
\begin{cases}
\frac{\partial \mathbf{u}}{\partial t} + f \unitvec \times \mathbf{u} - \nabla h = \mathbf{0} \\
\frac{\partial h}{\partial t} + \mathcal{H} \nabla \cdot \mathbf{u} = 0
\end{cases}
\end{align}
where $f = 2 \Omega \cos(\theta) =  2 \Omega z$ is the Coriolis parameter (note that the operator on the coefficients vector for multiplication by this parameter would be $F = 2 \Omega (J^z_\nabla)^T$).

Using backward Euler:
\begin{align}
\mathbf{u}_{n+1} &= \mathbf{u}_{n} + \Delta t \: (\nabla h_{n+1} - f \unitvec \times \mathbf{u}_{n+1}) \\
h_{n+1} &= h_n - \Delta t \: \mathcal{H} \nabla \cdot \mathbf{u}_{n+1}
\end{align}

Then using the operators on the coefficient vectors, the timestepping method is: 
\begin{align}
\mathbf{u}^c_{n+1} &= \mathbf{u}^c_{n} + \Delta t \: (G \mathbf{h}^c_{n+1} - F K \mathbf{u}^c_{n+1}) \\
\mathbf{h}^c_{n+1} &= \mathbf{h}^c_n - \Delta t \: \mathcal{H} D \mathbf{u}^c_{n+1}
\end{align}
$\iff$
\begin{align}
\mathbf{u}^c_{n+1} &= (I + \Delta t F K + \Delta t^2 \mathcal{H} G D)^{-1} \: (\mathbf{u}^c_{n} + \Delta t \: G \mathbf{h}^c_n) \\
\mathbf{h}^c_{n+1} &= \mathbf{h}^c_n - \Delta t \: \mathcal{H} D \mathbf{u}^c_{n+1}
\end{align}
where $F, K, G, D$ are all operators corresponding to multiplying by the Coriolis parameter, outward normal vector cross product, gradient and divergence operations respectively. It is important to note that these can be easily constructed due to the relationships between the vector spherical harmonics and scalar spherical harmonics we have outlined, and have a sparse Block-Banded structure.

These ideas will also be used to gain similar relationships for the circle arc, half disk, and spherical triangle orthogonal polynomials.

\section{Orthogonal polynomials on the circle arc}

As a stepping stone to working on spherical triangles, we will try to construct orthogonal polynomials on a circle arc, given by $x \in [h, 1]$, $y = (1-x^2)^\frac{1}{2}$. We will use the Lanczos algorithm as implemented in the ApproxFun package \cite{ApproxFun} to obtain 1D OPs $\{T^h_n\}$ (orthogonal wrt $w(x) = (1-x^2)^{-\frac{1}{2}}$) and $\{U^h_n\}$ (orthogonal wrt $w(x) = (1-x^2)^{\frac{1}{2}}$) on the interval $[h,1]$, and create our 2D OPs according to \cite{huybrechs2010fourier}:
\begin{align}
\bigP_0(x,y) := T^h_0(x) \equiv T_0,
\quad \quad 
\bigP_n(x,y) := \begin{bmatrix}
		T^h_n(x) \\
		y \: U^h_{n-1}(x)
	\end{bmatrix} \in \R^2 \quad \forall n \in \N,
\quad \quad 
\bigP := \begin{bmatrix}
		\bigP_0 \\
		\hline
		\bigP_1 \\
		\hline
		\bigP_2 \\
		\vdots \\
	\end{bmatrix}.
\end{align}

%%%%%%%%%%%%%%%%%%%%%%%%

\section{Orthogonal Polynomials on the disk-slice (2D)}

Sparse spectral methods for solving partial differential equations have been derived in recent years using hierarchies of classical orthogonal polynomials on intervals, disks, and triangles. In the work in this section we extend this methodology to a hierarchy of non-classical orthogonal polynomials on disk slices and trapeziums. This builds on the observation that sparsity is guaranteed due to the boundary being defined by an algebraic curve, and that the entries of partial differential operators can be determined using formulae in terms of (non-classical) univariate orthogonal polynomials. We apply the framework to solving the Poisson, variable coefficient Helmholtz, and Biharmonic equations. For now, we have focused on constant Dirichlet boundary conditions, as well as zero Dirichlet and Neumann boundary conditions, with other types of boundary conditions requiring future work.

Here, we have developed sparse spectral methods for solving linear partial differential equations on a special class of geometries that includes disk slices and trapeziums. More precisely, we consider the solution of partial differential equations on the domain
\begin{align}
	\Omega := \{(x,y) \in \R^2 \quad | \quad \alpha < x < \beta, \: \gamma \rho(x) < y < \delta \rho(x)\} \label{eqn:domaindef}
\end{align}
where either of the following conditions hold:

@@

For simplicity of presentation we focus on the disk-slice, where $\rho(x) = \sqrt{1-x^2}$, $(\alpha,\beta) \subset (0, 1)$, and $(\gamma, \delta)  = (-1,1)$, and discuss an extension to other the end-disk-slice in the appendix (note we omit detailed discussion  on the extension to trapezium in this document).

@@

We can express the three-term recurrences associated with $\genjac_n^{(a,b,c)}$ and $\jac_n^{(d,c)}$ as 
\begin{align}
	x \genjac_n^{(a,b,c)}(x) &= \beta_n^{(a,b,c)} \genjac_{n+1}^{(a,b,c)}(x) + \alpha_n^{(a,b,c)} \genjac_n^{(a,b,c)}(x) + \beta_{n-1}^{(a,b,c)} \genjac_{n-1}^{(a,b,c)}(x) \label{eqn:Hrecurrence} \\
	y \jac_n^{(d,c)}(y) &= \delta_n^{(d,c)} \jac_{n+1}^{(d,c)}(y) + \gamma_{n}^{(d,c)} \jac_{n}^{(d,c)}(y) + \delta_{n-1}^{(d,c)} \jac_{n-1}^{(d,c)}(y). \label{eqn:Precurrence}
\end{align}
Of course, for the disk-slice case, we have that $c = d$ and $\gamma_{n}^{(c,c)} = 0 \quad \forall n=0,1,2,\dots$. We can use (\ref{eqn:Hrecurrence}) and (\ref{eqn:Precurrence}) to determine the 2D recurrences for $\hdopnkabcd(x,y)$. Importantly, we can deduce sparsity in the recurrence relationships:

@@

%
\subsection{Sparse partial differential operators}\label{Section:PDOs}

@@

%\begin{proof}
%First, note that:
%\begin{align}
%	\genjacw^{(a,b,c) \: \prime}(x) &= - a \: \genjacw^{(a-1,b,c)}(x) + b \: \genjacw^{(a,b-1,c)}(x) + c \: \rho(x) \: \rho'(x) \:\genjacw^{(a,b,c-2)}(x), \label{eqn:derivativeofweightgenjac} \\
%	\jacw^{(c) \: \prime}(y) &= - 2c \: y \: \jacw^{(c-1)}(y), \label{eqn:derivativeofweightjac} \\
%	\rho(x) \: \rho'(x) &= -x. \label{eqn:rhoderivative}
%\end{align}
%
%We proceed with the case for the operator $D_y^{(a,b,c)}$ for partial differentiation by $y$. Since $\{\hdop^{(a,b,c+1)}_{m,j}\}$ for $m = 0,\dots,n-1$, $j = 0,\dots,m$ is an orthogonal basis for any degree $n-1$ polynomial, we can expand $\pddy \hdopnkabc = \sum_{m=0}^{n-1} \sum_{j=0}^m c_{m,j}^y \: \hdop^{(a, b, c+1)}_{m,j}$. The coefficients of the expansion are then the entries of the relevant operator matrix. We can use an integration-by-parts argument to show that the only non-zero coefficient of this expansion is when $m = n-1$, $j = k-1$. First, note that
%\begin{align*}
%	c_{m,j}^y = {\ip<\pddy \hdopnkabc, \hdop^{(a, b, c+1)}_{m,j} >_{W^{(a, b, c+1)}}}{\norm{\hdop^{(a, b, c+1)}_{m,j}}^{-2}_{W^{(a, b+1)}}}.
%\end{align*}
%Then, using the change of variable $t = \frac{y}{\rho(x)}$, we have that
%\begin{align*}
%	&\ip<\pddy \hdopnkabc, \hdop^{(a, b, c+1)}_{m,j} >_{W^{(a, b, c+1)}} \\
%	&= \iint_\Omega \: \Big[ \genjacnmk^{(a,b,2c+2k+1)}(x) \: \rho(x)^{k-1} \: \jac_k^{(c,c) \: \prime}\fpr(\frac{y}{\rho(x)}) \\
%	& \quad \quad \quad \quad \quad \quad \cdot \: \genjacmmj^{(a,b,2c+2j+3)}(x) \: \rho(x)^{j} \: \jac_j^{(c+1,c+1)}\fpr(\frac{y}{\rho(x)}) \Big] \: \D y \: \D x \\
%	&= \normgenjac^{(a, b, 2c+2k+1)} \: \ip< \genjacnmk^{(a,b,2c+2k+1)}, \rho(x)^{j-k+1} \: \genjacmmj^{(a,b,2c+2j+3)} >_{\genjacw^{(a,b,2c+2k+1)}} \\
%	&\quad \quad \quad \quad \quad \quad \cdot \: \normjac^{(c+1)} \: \ip< \jac_k^{(c,c) \: \prime}, \: \jac_{j}^{(c+1,c+1)} >_{\jacw^{(c+1)}}
%\end{align*}
%Now, using (\ref{eqn:derivativeofweightjac}), integration-by-parts, and noting that the weight $\jacw^{(c)}$ is a polynomial of degree $2c$ and vanishes at the limits of the integral for positive parameter $c$, we have that
%\begin{align*}
%	\normjac^{(c+1)} \: \ip< \jac_k^{(c,c) \: \prime}, \: \jac_{j}^{(c+1,c+1)} >_{\jacw^{(c+1)}} &= \int_{\gamma}^\delta \: \jac_k^{(c,c) \: \prime}(y) \: \jac_{j}^{(c+1,c+1)}(y) \: \jacw^{(c+1)}(y) \: \D y \\
%	&= - \int_{-1}^1 \: \jac_k^{(c,c)}(y) \: \ddy [\jacw^{(c+1)}(y) \: \jac_{j}^{(c+1,c+1)}(y)] \: \D y \\
%	&= - \int_{-1}^1 \: \jac_k^{(c,c)} \: [ \jac_{j}^{(c+1,c+1) \: \prime} \: \jacw^{(c+1)} - 2 c \: y \: \jac_{j}^{(c+1,c+1)} \:\jacw^{(c)} ] \: \D y \\
%	&= - \: \normjac^{(c)} \: \ip< \jac_k^{(c,c)}, \: \jacw^{(1)} \: \jac_{j}^{(c+1,c+1) \: \prime} -2c \:y \: \jac_{j}^{(c+1,c+1)} >_{\jacw^{(c)}}
%\end{align*}
%which is zero for $j < k-1$ by orthogonality. Further, when $j = k-1$, we have that
%\begin{align*}
%	&\normgenjac^{(a, b, 2c+2k+1)} \: \ip< \genjacnmk^{(a,b,2c+2k+1)}, \rho(x)^{j-k+1} \: \genjacmmj^{(a,b,2c+2j+3)} >_{\genjacw^{(a,b,2c+2k+1)}} \\
%	&= \normgenjac^{(a, b, 2c+2k+1)} \: \ip< \genjacnmk^{(a,b,2c+2k+1)}, \genjacmmj^{(a,b,2c+2k+1)} >_{\genjacw^{(a,b,2c+2k+1)}} \\
%	&= \normgenjac^{(a, b, 2c+2k+1)} \: \delta_{n,m+1},
%\end{align*}
%showing that the only possible non-zero coefficient is when $m=n-1, j=k-1$. Finally,
%\begin{align*}
%	c_{n-1,k-1}^y &= \ip< \jac_k^{(c,c) \: \prime}, \: \jac_{k-1}^{(c+1,c+1)} >_{\jacw^{(c+1)}}.
%\end{align*}
%
%We next consider the case for the operator $D_x^{(a,b,c)}$ for partial differentiation by $x$. Since $\{\hdop^{(a+1, b+1, c+1)}_{m,j}\}$ for $m = 0,\dots,n-1$, $j = 0,\dots,m$ is an orthogonal basis for any degree $n-1$ polynomial, we can expand $\pddx \hdopnkabc = \sum_{m=0}^{n-1} \sum_{j=0}^m c_{m,j}^x \: \hdop^{(a+1, b+1, c+1)}_{m,j}$. The coefficients of the expansion are then the entries of the relevant operator matrix. As before, we can use an integration-by-parts argument to show that the only non-zero coefficients of this expansion are when $m = n-1, n-2, n-3$, $j = k, k-1, k-2$ and $0 \le j \le m$. First, note that
%\begin{align*}
